% Documentation for jigsaw-generator
% Copyright (C) 2015 Julian Gilbey
% This is free software, and you are welcome to redistribute it
% under certain conditions; see the COPYING file for details.

\documentclass{article}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{calc}
\usepackage{tikz}[2013/12/13 v3.0.0]
\usetikzlibrary{calc}
\usepackage{amsmath}
\usepackage{upquote}  % to get straight quote marks for code examples

\def\YAML{{\small YAML}}
\def\CMEP{{\small CMEP}}
\def\JSON{{\small JSON}}
\def\PDF{{\small PDF}}
\def\HTML{{\small HTML}}
\def\XML{{\small XML}}
\def\MathML{Math{\small ML}}
\def\MacOSX{Mac\,{\small OS\,X}}
\def\GNU{{\small GNU}}
\def\WYSIWYG{{\small WYSIWYG}}
\def\PGF{{\small PGF}}
\def\TikZ{Ti\emph{k}Z}
\def\CSS{{\small CSS}}
\def\CTAN{{\small CTAN}}
\def\TUG{{\small TUG}}

\usepackage{shortvrb}
\MakeShortVerb{\|}

\def\gobble#1{}

%%%%%%%%%%%%%% The following are macros required for typesetting the
%%%%%%%%%%%%%% jigsaw example.  They are reduced from the full macros
%%%%%%%%%%%%%% which come with the jigsaw software.
\tikzset{regular/.style={sloped, above}}

\newlength\sidelength % side length
\newlength\trad % triangle circumradius
\newcommand{\setshapesize}[1]{%
 \pgfmathsetlength\sidelength{#1*72.27/2.54}
 \pgfmathsetlength\trad{(#1*72.27/2.54) / sqrt(3)}
}

\newcommand{\makestyles}[3]{%
    % split #3 into {style}{text}
    \makestylesa#1#2#3
}
\newcommand{\makestylesa}[4]{%
  \def#1{#3}
  \def#2{#4}
}

\newcommand{\tiltedtriangle}[7]{%
  \coordinate (a) at ($ #1 +({#2-150}:\trad) $);
  \coordinate (b) at ($ #1 +({#2-30}:\trad) $);
  \coordinate (c) at ($ #1 +({#2+90}:\trad) $);

  \makestyles\stylea\contenta{#3}
  \makestyles\styleb\contentb{#4}
  \makestyles\stylec\contentc{#5}

  \draw[allow upside down]
    (a) -- node[\stylea] {\contenta}
    (b) -- node[\styleb] {\contentb}
    (c) -- node[\stylec] {\contentc}
    cycle;
  % If card number is empty, don't draw circle around number
  \def\templabel{#6}
  \ifx\templabel\empty
  \else
    \node[circle,draw,thin,inner sep=1pt,rotate=#7] at ($ #1 $) {#6};
  \fi
}

\newcommand{\uptriangle}[6]{%
  \tiltedtriangle{#1}{0}{#2}{#3}{#4}{#5}{#6}
}
\newcommand{\downtriangle}[6]{%
  \tiltedtriangle{#1}{180}{#2}{#3}{#4}{#5}{#6}
}
\newcommand{\lefttriangle}[6]{%
  \tiltedtriangle{#1}{90}{#2}{#3}{#4}{#5}{#6}
}
\newcommand{\righttriangle}[6]{%
  \tiltedtriangle{#1}{270}{#2}{#3}{#4}{#5}{#6}
}

%%%%%%%%%%%%%% end jigsaw macros


\begin{document}

\title{Creating (mathematical) jigsaw puzzles using jigsaw-generator}
\author{Julian Gilbey}
\date{December 2015}
\maketitle

\begin{abstract}
  \noindent
  The jigsaw puzzles described here are constructed from shapes such
  as triangles, squares and so on, with questions and answers written
  along their edges.  The aim is to match them up correctly.  Related
  puzzle varieties are card sorts and dominoes.  This package provides
  a \TeX- and Python-based system designed to author such puzzles.
  The input is a text-based \YAML\ file; the output can include both
  printable \PDF s for cutting up and Markdown files for potential
  conversion to \HTML.
\end{abstract}

\tableofcontents

\section{Introduction}

For a general background to this software and its history, see
my \TUG boat article~\cite{JGtugboat}.

This documentation describes the usage of the software; some details
have changed since the \TUG boat article, and so all of the relevant
material will be repeated here.

\subsection{Usage overview}

The most important files from a user's perspective are the data input
files.  These specify the type of puzzle (a hexagonal jigsaw or a card
sort, for example), any options for the jigsaw (such as whether to
number the cards or not) and the text to appear on the puzzle.
Sensible defaults are provided for all of the options if they are not
specified.  These input files are written in \YAML.

For example, here is a data file for creating a small hexagonal puzzle:

\begin{verbatim}
type: smallhexagon
title: An example puzzle
note: 'You will have to work out the missing number shown as `?'''

pairs:
  - ['$10^6\div10^3$', '$1000$']
  - ['$10^{23}\div10^{21}$', '$100$']
  - ['$9^4\times9^5$', '$9^9$']
  - - '$(3^2)^2$'
    - puzzletext: '?'
      solutiontext: '$81$'
  - ['$6\times6^2$', '$216$']
  - ['$2^4\div2^6$', '$\dfrac{1}{4}$']
edges:
  - '$-1$'
  - '$10$'
  - '$13^2\times13$'
  - '$39$'
  - '$13^3$'
  - '$108$'
\end{verbatim}

The resulting puzzle looks like this:

\begin{center}
  \tikzset{above/.default=2pt}
  
  \setshapesize{3.7}
  \begin{tikzpicture}[line width=1.5pt, rounded corners=0.7pt]
    \righttriangle{(60:\trad)}  {{regular}{\normalsize $1000$}}{{regular}{\normalsize $10^{23}\div10^{21}$}}{{regular}{\normalsize $108$}}{\scriptsize 1}{150} % {A1}{Q2}{E1}{1}{270}
    \lefttriangle {(0:\trad)}   {{regular}{\normalsize $13^3$}}{{regular}{\normalsize $100$}}{{regular}{\normalsize $9^4\times9^5$}}{\scriptsize 3}{-150} % {E2}{A2}{Q3}{2}{90}
    \righttriangle{(-60:\trad)} {{regular}{\normalsize $(3^2)^2$}}{{regular}{\normalsize $39$}}{{regular}{\normalsize $9^9$}}{\scriptsize 2}{150} % {Q4}{E3}{A3}{3}{270}
    \lefttriangle {(-120:\trad)}{{regular}{\normalsize $81$}}{{regular}{\normalsize $6\times6^2$}}{{regular}{\normalsize $13^2\times13$}}{\scriptsize \underline{6}}{-30} % {A4}{Q5}{E4}{4}{90}
    \righttriangle{(180:\trad)} {{regular}{\normalsize $10$}}{{regular}{\normalsize $216$}}{{regular}{\normalsize $2^4\div2^6$}}{\scriptsize 5}{150} % {E5}{A5}{Q6}{5}{270}
    \lefttriangle {(120:\trad)} {{regular}{\normalsize $10^6\div10^3$}}{{regular}{\normalsize $-1$}}{{regular}{\normalsize $\frac{1}{4}$}}{\scriptsize 4}{-150} % {Q1}{E6}{A6}{6}{90}
  \end{tikzpicture}
\end{center}

If the puzzle is stored in the file |hexpuzzle.yaml|, then the
command:
\begin{verbatim}
jigsaw-generate hexpuzzle
\end{verbatim}
will produce several \PDF\ files: |hexpuzzle-puzzle.pdf|
(containing the puzzle), |hexpuzzle-solution.pdf| (the solution)
and |hexpuzzle-table.pdf| (with the puzzle content laid out in a
tabular form), along with Markdown versions |hexpuzzle-puzzle.md|
and |hexpu|zzle-solution.md+ containing the puzzle and solution
respectively; these can be used to create \HTML\ versions of the
puzzle via pandoc or some such system.

In the following sections, the data files are described in detail, and
then the other supporting files (configuration files, templates and so
on) and command line options are described.

In many cases, it is adequate to simply copy and edit the example
input files, which will be found in
\nolinkurl{/usr/local/share/doc/jigsaw-generator/examples} or
\nolinkurl{/usr/share/doc/jigsaw-generator/examples}, depending on
your specific installation of the package.

\section{Data input files}

The data input files consist of layout information (the type of
puzzle, whether to produce a solution and so forth) and content
information.  Being a \YAML\ mapping, these can appear in any order in
the data file; will will describe them in these two separate sections
for simplicity.

\subsection{Available layouts}

There are four jigsaw puzzle layouts currently available:
\begin{itemize}
\item smallhexagon, which consists of 6 triangles as shown above;
\item hexagon, which consists of 24 triangles;
\item triangle, which consists of 16 smaller triangles;
\item parquet, which consists of 4 squares and 8 triangles.
\end{itemize}

There are also three card sort layouts currently available:
\begin{itemize}
\item cards, which consists of an arbitrary number of rectangular
  cards; no solution is produced, and this is intended for
  classification activities or suchlike, where there is no ``correct''
  answer expected; these cards are not shuffled by default;
\item cardsort, which is like cards, except that there is a correct
  order, so the cards are shuffled and a solution is produced;
\item dominoes, which consists of an arbitrary number of domino cards;
  a solution is produced.
\end{itemize}

More details are given in the following sections.

\subsection{Layout information}

Each of these options must be followed by a colon and a space,
followed by the option value, as demonstrated in the example above.
All of these options are optional unless stated otherwise.  The most
common options are given first in each of the following three lists.

\subsubsection*{Options applicable to all puzzle types}

\begin{itemize}
\item |type:| (required) The type of jigsaw being created.  The
  current recognised standard types are those listed above, namely
  |smallhexagon|, and so on.
\item |title:| The title of the activity, printed at
  the top of the page.
\item |note:| A note printed at the top of the page.
\item |produceSolution:| Whether to produce a solution \PDF\ and
  Markdown file; the default is |true| except for cards, where
  the default is |false|.
\item |numberCards:| Do we number the cards for ease of
  identification?  (Default: |true|)
\item |textSize:| How large the text should be on the cards; this
  is a number from 0 to 9, with 0 being the smallest.  (Default:~5)
\item |makepdf:| Whether to produce \PDF\ output files.  (Default:
  |True|)
\item |makemd:| Whether to produce Markdown output files.  (Default:
  |True|)
\item |latex:| Which \LaTeX\ engine to use to produce the \PDF\
  files.  (Default: |pdflatex|)
\item |texfilter:| If specified, this is a filter to pass the \LaTeX\
  files through prior to running \LaTeX.  See the later section on
  filters (\ref{sec:filters}) for more information on filters.
\item |mdfilter:| If specified, this is a filter to pass the Markdown
  files through to produce the final Markdown files.
\item |clean:| If true, then intermediate files will be deleted.
  This includes the \LaTeX\ files and related aux files and so on.
  (Default: |true|)
\item |puzzleHeaderTeX:| This specifies the header file to use to
  produce the \LaTeX\ puzzle file.  See the later section on templates
  (\ref{sec:templates}) for more information on header and template
  files.
\item |puzzleTemplateTeX:| This specifies the template file to
  use to produce the \LaTeX\ puzzle file.
\item |solutionHeaderTeX:| This specifies the header file to
  use to produce the \LaTeX\ solution file.
\item |solutionTemplateTeX:| This specifies the template file to
  use to produce the \LaTeX\ solution file.
\item |tableHeaderTeX:| This specifies the header file to
  use to produce the \LaTeX\ table file.
\item |tableTemplateTeX:| This specifies the template file to
  use to produce the \LaTeX\ table file.
\item |puzzleHeaderMarkdown:| This specifies the header file to
  use to produce the Markdown puzzle file.
\item |puzzleTemplateMarkdown:| This specifies the template file to
  use to produce the Markdown puzzle file.
\item |solutionHeaderMarkdown:| This specifies the header file to
  use to produce the Markdown solution file.
\item |solutionTemplateMarkdown:| This specifies the template file to
  use to produce the Markdown solution file.
\item |hiddennote:| If any entries are hidden, the note to include at
  the top of the \PDF\ solution output file.  (Default:
  |'|\texttt{Entries that are hidden in the puzzle are highlighted in
    yellow.}|'|)
\item |hiddennotemd:| If any entries are hidden, the note to include
  at the top of the Markdown solution output file.  (Default:
  |'|\texttt{Entries that are hidden in the puzzle are indicated with
    (*).}|'|)
\item |hiddennotetable:| If any entries are hidden, the note to
  include at the top of the \PDF\ table output file.  (Default:
  |'|\texttt{Entries that are hidden in the puzzle are indicated with
    (*).}|'|)
\end{itemize}

\subsubsection*{Options applicable to card sorts and dominoes only}

\begin{itemize}
\item |rows:| (required) How many rows of cards to produce per
  page.
\item |columns:| (required) How many columns of cards to produce
  per page.
\item |shuffleCards:| (card sorts only) Are the cards shuffled?
  This is by default |true| for cardsort, and |false| for
  cards.
\item |shufflePairs:| (dominoes only) Are the order of the pairs
  shuffled from the order given in the data file?  This is by default
  |false|.
\item |flip:| (dominoes only) Are the order of the entries in
  each pair presented in a random order?  This is by default
  |false|.
\item |cardTitle:| The title to write on each card.
\item |cardTitleSize:| How large the title should be on the
  cards, a number from 0 to 9 (like the |textSize|).  By default,
  this is 1~less than the label size (see below).
\item |label:| The default label to write on each card.  This is
  like a subtitle, and appears at the bottom of the card.  (This can
  be useful if the cards are of two or more types, and the type of
  card is important.)  This is not used for dominoes.
\item |labelSize:| How large the labels should be on the cards,
  again a number from 0 to 9; by default, this is 2~less than the text
  size.  (This can be overridden for individual cards.)  This is not
  used for dominoes (though setting it will have an impact on the size
  of the title if |cardTitleSize| is not set).
\item |cardsep:| How much space to leave between
  adjacent cards.  By default this is 12pt.  It should be a \TeX\
  length.
\item |cardsepHorizontal:| How much space to leave between
  horizontally adjacent cards.  By default, this is |cardsep|.
\item |cardsepVertical:| How much space to leave between
  vertically adjacent cards.  By default, this is |cardsep|.
\item |loop:| (dominoes only) Do we have a domino loop, with
  there being no start and end?  The default is |true|.
\item |start:| (dominoes only) If we do not have a loop, then the
  text to write on the start of the chain.  The default is |'Start'|.
\item |finish:| (dominoes only) If we do not have a loop, then the
  text to write on the end of the chain.  The default is |'Finish'|.
\item \textbf{WILL WANT SOMETHING HERE ABOUT \% DIVIDER ON DOMINOES}
\end{itemize}

\subsubsection*{Options applicable to jigsaws only}

\begin{itemize}
\item |shufflePairs:| Are the order of the pairs shuffled from
  the order given in the data file?  This is by default |false|.
\item |flip:| Are the order of the entries in each pair presented
  in a random order?  This is by default |true|.
\item |shuffleEdges:| Are the order of the edges shuffled from
  the order given in the data file?  This is by default |false|.
\end{itemize}

\subsection{Card content}

The content of the cards is given in a final one or two \YAML\
key--value pairs, depending on the type of material being produced.
The keys are |cards| (for card sorts), |pairs| (for dominoes and
jigsaws) and |edges| (for jigsaws).

\section{Old stuff dumped here still to tidy}

The puzzle templates are \LaTeX\ files with templating marks.  For the
templates I have created, I have used \PGF/\TikZ\ for the graphics and
to place the text items.  My initial templates were written when I was
using version 2.10 of \PGF/\TikZ, as distributed with \TeX~Live 2013.
However, it turned out that I ran into a bug related to the placement
of text along cyclic paths, which has been fixed in version 3.0.0
(distributed in \TeX~Live 2014).  It is therefore necessary to have an
up-to-date \TeX~Live distribution (or at least an up-to-date \PGF) for
these templates to work correctly.  Alternative templates can be
written if this is desired.  For example, someone might prefer to use
Asymptote or another graphical package, or they may wish to modify the
existing templates in various ways.

There is also a template description file (again written in \YAML) for
each puzzle type: this specifies various parameters required to create
the puzzle.

Both of these template types (the puzzle templates and the description
files) are described in detail in the software documentation.

I wrote the program itself in Python.  This was for a few reasons.
Firstly, it is a well-known, popular language, so if other people wish
to become involved in developing this software, it will be relatively
easy for them to do so.  Secondly, the Python interpreter is easy to
install on the major platforms people will be using.  Furthermore,
since the software is written in pure Python, it does not require
compilation, making things a little simpler to install.  Finally, it
provided me with an opportunity to learn more about this language: as
I have learnt more about Python, I have improved my code and made it
more idiomatic.  The code is currently written in Python~3.x; if I
have time and people express an interest, I will endeavour to make it
compatible with Python~2.7.

To create the jigsaws, the jigsaw generator program is run over the
data file.  It reads this file along with the relevant template
description file and puzzle template files.  It then fills in the
puzzle templates with the puzzle data to create \LaTeX\ files (and
Markdown files too, if requested).  The \LaTeX\ files are then
processed with pdf\LaTeX\ (or some other \LaTeX\ variant) to create
\PDF s for printing.

\subsection{An example data file}

Figure~\ref{fig:smallhex-example} shows the \YAML\ puzzle file which
was used to create the example shown in
figure~\ref{fig:jigsaw-complete} above (with some small
modifications).

The file begins with some metadata:
\begin{itemize}
\item the type of the puzzle: in this case, it is a `smallhexagon'
  puzzle, which consists of six triangles;
\item the title of the puzzle, which is optional, and
\item an optional note, which is printed above the puzzle.
\end{itemize}

The existing jigsaw types at the time of writing are \texttt{hexagon},
\texttt{smallhexagon}, \texttt{triangle} and \texttt{parquet}; there are a few more in the
pipeline, too.  It is also possible to write one's own jigsaw types,
as long as they consist of equilateral triangles and squares.  (To
write templates using different shapes would require extensions to the
software itself; I may do this in the future.)  In addition, there are
three more types, \texttt{cards}, \texttt{cardsort} and \texttt{dominoes}, which are described
briefly later.

The next section of the file specifies the data.  There are two parts
here: the pairs data, which lists question and answer pairs, and the
edges data, which specifies the text to appear on the edges in an
anticlockwise direction.  Each pair is a sequence of two items, the
question and the answer, whereas each edge consists of just a single
item.

Each item is usually just a single string.  However, there is a
possibility of `hiding' text in the puzzle.  This means either leaving
a blank where the text should appear or writing something else in its
place.  The fourth question--answer pair in the above example
illustrates this.  This pair is written using \YAML\ `block style' for
clarity, that is, the question and answer are written on separate
equally-indented lines.  The question is just a plain string
(|'$(3^2)^2$'|), whereas the answer is a mapping with two
entries: the \texttt{puzzletext} appears in the puzzle, while the
\texttt{solutiontext} appears in the solution.  This feature could be
used as in the example shown, or it could be used to introduce
deliberate mistakes in the puzzle, increasing the level of difficulty
for the students.  There is also an alternative notation available to
simply hide the text in the puzzle, which is described in the
documentation.

If a puzzle does include such hidden text, then the solution
highlights these occurrences; in the default templates, these are
shown with a yellow background.

It is also possible to include images in the items or to change the
text size of individual items or all items.  The details are described
in the documentation.

\subsubsection{Some notes on \texorpdfstring{\YAML}{YAML} syntax}

For a full, precise description of \YAML\ syntax, see the \YAML\
Specification~\cite{YAML}.  What follows is a very brief summary of
some of the basic parts of the specification which should be
sufficient for most people's needs when using this software.

Although \YAML\ allows unquoted strings in general, there are a number
of restrictions on what is permitted in them.  For this reason, it may
be simplest to single-quote all value strings.  (\YAML\ also offers
double-quoted strings, but these interpolate backslash-escapes; this
is probably undesirable in this context, since many \TeX\ expressions
include backslashes.)

Within a single-quoted string, a single quote is written as a doubled
quote mark (|''|), hence the three quote marks in a row at the
end of the note field in the above example (two to indicate a quote,
and the third to end the string).

For collections (`sequences' and `mappings' in \YAML's terminology,
each of which consists of a number of `entries'), \YAML\ allows two
different notations: either a flow style, which is similar to \JSON\
notation, or a block style.  With the flow style, the entries of a
sequence, such as a question--answer pair or the list of edges,
are enclosed in square brackets and separated by commas; for the block
style, each entry appears on a new line preceded by a
vertically-aligned hyphen.

Similarly, for a mapping the entries
consist of key--value pairs, with the key and value separated by a
colon.  They can be written either using a flow style as a
comma-separated list enclosed in braces, or with a block style by
writing each key--value pair on an identically-indented new line.

Both of the sequence styles appear in the example here, though only
the block style is used for mappings.  Note also that the top-level
structure of the file is itself a mapping.  This means that the
entries (type, title, pairs, and so on) can appear in any order.
However, for the benefit of the human reader, it is wise to maintain a
meaningful order to these entries.

\subsection{Card sorts and dominoes}

As mentioned above, this software also offers some other types of
activities: card sorts and dominoes.

A domino puzzle is just a collection of question--answer pairs
which are laid out on a series of dominoes.  Each domino consists of
an answer and a new question.  The dominoes can then be laid out to
create a complete chain (beginning with `Start' and ending with
`Finish') or loop (if `Start' and `Finish' are not present), with each
question matching its corresponding answer.  The data file used to
create this is very similar to the example shown above, only the type is
now \texttt{dominoes}.  Within the data file, it is also possible to specify
various options such as how many dominoes should appear on a page and
whether the cards should form a loop or chain; these are described in
more detail in the documentation.

A card sort is simply a collection of cards which are to be sorted in
some way.  For the \texttt{cardsort} type, the aim is to sort the cards into
the correct order, and so the cards are shuffled for the puzzle and a
solution is also produced by default.  For the \texttt{cards} type, on the
other hand, there is no canonical order (for example: ``Arrange these
cards into order of importance to you'' in a politics
lesson), and so no solution is produced, nor are the cards shuffled by
default.  Again, the data file is very similar in structure, and
details can be found in the documentation.

Both cards and dominoes offer the possibility of having some form of
title on the cards, and cards have a further option of having labels
on individual cards.  (This was introduced into the software when we
wanted to create an activity which had different categories of cards;
the categories were then written on the cards.)  Additional options
are discussed in the documentation.

\subsection{Markdown output}

As explained in section~\ref{sec:desire-tex}, our requirements include
the need to output a Markdown version of cards data for inclusion into
our build system; from there, it is translated into \HTML.  Our
current system requires each card to be embedded in an \HTML\
|<div>| element.  The cards are then displayed in an appropriate
way using some simple \CSS.  Since |pandoc| passes any \HTML\
|<div>| elements in a Markdown file to the \HTML\ output
unchanged, we simply need a Markdown file with the |<div>|
elements already present for our card sorts.  I have therefore created
a Markdown template file which places the card content within these
elements.

For other needs, it is perfectly straightforward to create alternative
Markdown templates, which could then be converted into the required
\HTML.  For example, with the appropriate JavaScript and supporting
\CSS, it would be entirely feasible to create an interactive version
of the card sort or jigsaw from the same data file.  While we have not
yet done this, it would be a very interesting next step.

\section{Installing the package}

The software can be downloaded from GitHub; the repository is
\url{https://github.com/juliangilbey/jigsaw-generator}.  At the time
of writing, the simplest way to obtain it is to use git to clone the
repository.

The software has been tested on \GNU/Linux and \MacOSX\ operating
systems.

\smallskip
\section*{Acknowledgements}

I thank those who offered very useful ideas and feedback at the \TUG\
conference; these have helped me to solve some of the thorny issues I
had been facing.

\smallskip
\bibliographystyle{plain}
\bibliography{jigsawrefs}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
